---
title: "R Notebook"
output: html_notebook
---

Example set of steps to download data from the ALA and then filter them based on QA/QC fields.

Adapted from code used for:
Cassis, G., Laffan, S.W. and Ebach, M.C. (2017) Biodiversity and bioregionalisation perspectives on the historical biogeography of Australia. In Ebach, M.C (ed) Handbook of Australasian Biogeography, ch 1, pp 1-16.
http://www.crcnetbase.com/doi/abs/10.1201/9781315373096-2

That code was in turn adapted from code written by Hugh Burley, who adapted code from Nick dos Remedios at the ALA.

Later additions have cribbed from https://atlasoflivingaustralia.github.io/ALA4R/articles/ALA4R.html and https://atlasoflivingaustralia.github.io/galah/articles/galah.html 

```{r setup}
library (galah)
library (stringr)
library (rgdal)
library(sf)
library(maps)
library(mapdata)
library(dplyr)
library(tidyr)
library(ggplot2)
#  needs additional system level libs for this to work
# library (CoordinateCleaner)
```

Configuration section, including filters.

The location filter is a simple rectangular bounding box.  
The data quality filter uses the default ALA profile to strip out low quality records.  This gets most of them, but not all.


```{r}

#  An email address is needed to download
#  data from the ALA - you need to provide one
email_address <- ""
if (email_address == "") {message ("EMAIL ADDRESS NOT SET")}

#  make sure we use a local cache on the EcoCloud
if (Sys.getenv("USER") == "jovyan") {
  dir.create("~/ala_cache", showWarnings=FALSE)
  ala_config(cache_directory="~/ala_cache")
}
ala_config (caching=TRUE)
# ala_config (caching="off")
# ala_config(warn_on_empty=TRUE)

#  search spatialreference.org for coordinate system definitions
in_prj  <- paste0("+init=epsg:4326") # WGS84 in decimal degrees
out_prj <- paste0("+init=epsg:3577") # Albers equal area projection for Australia

failed <- c()

#  Define a bounding box.
#  The template approach is awkward and inefficient
#  Need a better way, perhaps using sprintf
bbox_template <- "POLYGON((e s, e n, w n, w s, e s))"

#  bbox for the globe
bb <- list (e=180, w=-180, n=90, s=-90)
#  rough bounding box for Australia as WKT format, edit as appropriate
bb <- list (e=154, w=112.9, n=-9, s=-43.74)
#  approx NSW for tests
# bb <- list (e=153, w=140, n=-28, s=-37)

bbox = bbox_template
for (key in names(bb)) {
  bbox <- gsub (key, bb[key], bbox)
}

location_filter <- select_locations(query = bbox)


#  these are the colnames we want in the output
subset_colnames <- c(
  "id",     "x",     "y",       "scientificName",
  "family", "genus", "species", "rank", 
  "coordinateUncertaintyInMetres",
  "decimalLongitude", "decimalLatitude"
)

column_subset <- select_columns(group = "basic")

selection_filters <- select_filters(
  #  throws an error on ecocloud
  # basisOfRecord = "HumanObservation",  
  profile = "ALA"
)


```

```{r}

#  Edit as needed.  Some can be case sensitive, 
#  so check the ALA website to be certain.
#  We also need a kingdom for Corymbia because there
#  is a suppressed homonym from the Coleoptera
taxon_list <- list (
  "Corymbia"   
    = list (genus = "Corymbia", kingdom="Plantae")
)

taxon_list
```

Create the target directories and download the data.
Note that the directories will be in your current working directory.  

```{r}



for (dir in names(taxon_list)) {
  if (!dir.exists(dir)) {
    print (paste("Creating directory", dir))
    dir.create(dir, showWarnings = TRUE)
  }

  collated_df = data.frame()
  
  search_term = taxon_list[[dir]]

  print (paste0 ("Search term is ", paste (search_term)))

  occ <- ala_occurrences(
    taxa      = select_taxa(query = search_term),
    filters   = selection_filters,
    locations = location_filter,
    columns   = column_subset
  )

  head(occ)
  print (dim (occ))

  coords_ll = occ[c("decimalLongitude", "decimalLatitude")] %>% 
    as.data.frame %>% 
    sf::st_as_sf(coords = c(1,2), crs=in_prj)

  #  no doubt this could be dplyrd
  coords_albers = st_transform(coords_ll, crs = st_crs(out_prj))
  coords_albers = st_coordinates(coords_albers)
  colnames(coords_albers) = c("x","y")
  occ = cbind(occ, coords_albers)

  write.csv(occ, file=csv_fname)

  if (nrow(collated_df) == 0) {
    collated_df = subset
  } else {
    collated_df = rbind (collated_df, subset)
  }
  
  #  we could also save and re-use the occ df as an RDS file

}

collated_fname = paste0(dir, "/collated.csv")
write.csv (collated_df, collated_fname)

    
```
   

You can of course do things like plot the data.

Perhaps try to make something more aesthetically pleasing to the eye.  

```{r}

dir <- names(taxon_list)[1]
dir 

collated_fname <- paste0(dir, "/collated.csv")

df <- read.csv(collated_fname)

plot (df$x, df$y)



```

Calculate a sinple gridded species richness estimate.  

This code is directly adapted frm the ALA4R examples.  
https://atlasoflivingaustralia.github.io/ALA4R/articles/ALA4R.html


```{r}
xgridded <- collated_df %>%
    ## discard genus- and higher-level records
    dplyr::filter(rank %in%
                  c("species")) %>%

    ## bin into 0.5-degree bins
    mutate(longitude=round(longitude*2)/2, latitude=round(latitude*2)/2) %>%

    ## subset to vars of interest
    dplyr::select(longitude, latitude, scientificName) %>%

    ## take one row per cell per species (presence)
    distinct() %>%
  
    #  group by coordinates
    group_by (longitude, latitude) %>%

    ## calculate species richness
    mutate(richness=n()) %>%

    ## convert to wide format (sites by species)
    mutate(present=1) %>%
    do(tidyr::spread(data=., key=scientificName, value=present, fill=0)) %>%
    ungroup()

## where a species was not present, it will have NA: convert these to 0
sppcols <- setdiff(names(xgridded),
                   c("longitude", "latitude",
                     # "precipitationAnnual", "temperatureAnnualMaxMean",
                     "richness"))
xgridded <- xgridded %>% mutate_at(sppcols, function(z) ifelse(is.na(z), 0, z))

```
    
Plot by latitude and longitude, just because we can.

```{r}
library(ggplot2)
ggplot(xgridded, aes(latitude, richness)) + geom_point() + theme_bw()
ggplot(xgridded, aes(longitude, richness)) + geom_point() + theme_bw()


```

Plot a basic map of species richness.  People with skills in R plotting will be able to do a better job than this.

You could also connect it with phylogenetic data.  There are many tools to do so, but it is easily done in Biodiverse.  


```{r}
#  plot richness patterns

maps::map("worldHires", "Australia", xlim=c(105, 155), ylim=c(-45, -10), col="gray90", fill=TRUE)
palette (terrain.colors(max(xgridded$richness)/3))
with(xgridded, points(longitude, latitude, pch=21, 
                      col=richness, bg=richness,
                      cex=0.75))

```

    
Things to try:

See if you can edit the code above to use config variables,
i.e. instead of hard coding the error tolerance at 10000, 
why not set it in the config section? See the galah help pages for how to set filters.

You might also wish to plot the data sets before they are filtered to get a sense of where the errors are.  
